---
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r}
set.seed(2019)

particle_filter <- function(correction = TRUE, sd_emission = 1)
{
for (t in 1:T)
{
    for (m in 1:M)
    {
      
      # Sample from beliefs and weights in last time step
      
      if (t!=1)
      {
        samples[t,m] <<- sample(x=beliefs[t-1,],size=1,replace=TRUE,prob = weights[t-1,])
      } else {
        samples[t,m] <<- runif(n = 1,min = 0,max = 100)
      }
      
      # Predict the state in next step according to transition model
      beliefs[t,m] <<- rnorm(n = 1, mean = samples[t,m] + sample(x = -1:1, size = 1), sd = 1)
      
      # Calculate weights of how well the belief corresponds to the observation
      weights[t,m] <<- dnorm( x = observations[t], mean = beliefs[t,m], sd = sd_emission)
    }
 
    if(!correction)
    {
      weights[t,] <<- 1
    }
  
      # Normalize weights
    weights[t,] <<- weights[t,]/sum(weights[t,])
    
      # Calculate the expected value/prediction 
  predictions[t] <<- sum(beliefs[t,] * weights[t,])
}
  
  cat("Average error with: ", mean(abs(predictions-states)),"\n\n")
  
cat("Expected location in first step: \n", predictions[1], "\n")
cat("True location in first step: \n", states[1], "\n\n")
plot(density(beliefs[1,]))


cat("Expected location in step 33: \n", predictions[33], "\n")
cat("True location in step 33: \n", states[33], "\n\n")
plot(density(beliefs[33,]))

cat("Expected location in step 67: \n", predictions[67], "\n")
cat("True location in step 67: \n", states[67], "\n\n")
plot(density(beliefs[67,]))

cat("Expected location in last step: \n", predictions[T], "\n")
cat("True location in last step: \n", states[T], "\n\n")
plot(density(beliefs[T,]))
}
```

```{r}

#AMN
T<-100
mu_0<-50
Sigma_0<-10
R<-1
Q<-5

x<-vector(length=T)
z<-vector(length=T)
err<-vector(length=T)

for(t in 1:T){
  x[t]<-ifelse(t==1,rnorm(1,mu_0,Sigma_0),x[t-1]+1+rnorm(1,0,R))
  z[t]<-x[t]+rnorm(1,0,Q)
}

mu<-mu_0
Sigma<-Sigma_0*Sigma_0 # KF uses covariances
for(t in 2:2){
    pre_mu<-mu+1
  pre_Sigma<-Sigma+R*R # KF uses covariances
  K<-pre_Sigma/(pre_Sigma+Q*Q) # KF uses covariances
  mu<-pre_mu+K*(z[t]-pre_mu)
  Sigma<-(1-K)*pre_Sigma
  
  err[t]<-abs(x[t]-mu)
  
  #cat("t: ",t,", x_t: ",x[t],", E[x_t]: ",mu," , error: ",err[t],"\n")
  #flush.console()
}

mean(err[2:T])
sd(err[2:T])
```

```{r}
################################################################
# TASK 1
################################################################

T <- 100
M <- 100

# Beliefs represented by particles. These are obtained by sampling from the corrected beliefs in the last timestep and the transition distribution 
beliefs <- matrix(NA,T,M)

# Beliefs corrected by observations
samples <- matrix(NA,T,M)

states <- matrix(NA,T,1)
observations <- matrix(NA,T,1)
predictions <- matrix(NA,T,1)

# Weights corresponding to P(X_t|Z_t)
weights <- matrix(NA,T,M)

# Set initial state
states[1] <- runif(n = 1,min = 0,max = 100)

# Simulate the rest of the states
for (t in 2:T)
{
  states[t] <- rnorm(n = 1, mean = states[t-1] + sample(x = 0:2, size = 1), sd = 1)
}

# Simulate the observations
for (t in 1:T)
{
  observations[t] <- rnorm(n = 1, mean = (states[t] + sample(x = -1:1, size = 1)), sd = 1)
}

particle_filter()

```
```{r}

# TASK 2 (A)
sd_emission <-5

for (t in 1:T)
{
  observations[t] <- rnorm(n = 1, mean = states[t] + sample(x = -1:1, size = 1), sd = sd_emission)
}

particle_filter(sd_emission = sd_emission)

```

```{r}
# TASK 2 (B)
sd_emission <-50
for (t in 1:T)
{
  observations[t] <- rnorm(n = 1, mean = states[t] + sample(x = -1:1, size = 1), sd = sd_emission)
}

particle_filter(sd_emission = sd_emission)

```

Since the observations become more uncertain, the predictions becomes worse. This is confirmed by the average error, with increases significantly.

```{r}
# TASK 3
particle_filter(FALSE)
```
When no corrections are made, the initial prediction becomes the average of the first particles. The rest of the predictions, no matter what the observations are, are made using the transition model. This results in that the prediction error is rather constant from the first prediction to the last, as shown in the output.

